---
title: "Thiolase substrate specificity prediction"
resource_files:
- www/grp1.png
- www/heptynoate.png
runtime: shiny
output:
  flexdashboard::flex_dashboard:
    css: bootstrap.css
    favicon: www/pNP_struct_white.png
    logo: www/pNP_struct_white.png
    social: menu
    theme: united
    vertical_layout: fill
---

```{r setup, include=FALSE}
# install.packages("BiocManager")
# options(repos = BiocManager::repositories())
# getOption("repos")
# setwd("~/Documents/University_of_Minnesota/Wackett_Lab/github/thiolase-machine-learning/shiny_app/")
# library("flexdashboard")
library("xtable")
library("dplyr")
library("data.table")
library("shiny")
library("Biostrings")
library("tidyverse")
library("knitr")
library("kableExtra")
library("readxl")
library("DECIPHER")
library("ranger")
library("muscle")
library("DT")
```

```{r global}
## Read in the training set
source("lib/extract_12angstrom_aas.R")
subkey <- read_excel("data/pNP_substrate_key.xlsx")
tab <- read_csv("data/20191218_all_cmpnds_avg_log_slopes_for_database.csv")
molec_fts <- read_csv("data/PC7_molecular_descriptors_clean.csv")
```

Home
=====================================
Column {.sidebar}
-----------------------------------------------------------------------
**Welcome to pNPred**
<br>
<br>
*Prediction tool for substrate specificity of OleA-type thiolase enzymes towards different p-nitrophenyl esters (pNPs)*
<br>
<br>
pNPred uses a supervised machine learning approach to predict pNP substrate specificity using thiolase amino acid sequences as inputs. 
<br>
<br>
This tool is maintained by [Dr. Larry Wackett’s lab](https://cbs.umn.edu/wackett-lab/home) at the University of Minnesota.

Column
-----------------------------------------------------------------------
Input a FASTA file containing nucleotide* or protein sequences (limit 100 sequences) to predict substrate specificity and functional class. 

```{r}
 # Input: Select a file ----
fileInput("file1", "Choose a FASTA File:**", multiple = FALSE)

splitLayout(
  cellWidths = 155,
  cellArgs = list(style = "padding: 5px"),
  actionButton("run", "Run Analysis", icon("paper-plane")),
  downloadButton("example", "Example FASTA", icon("file-download"))
)

# Horizontal line ----
tags$hr()

# Download button
downloadButton("downloadData", "Download", class = "download_this", width = '30%')

# Input: Select sequence type----
radioButtons("seqtype", "Sequence type",
             choices = c(Nucleotide = "nuc",
                         Protein = "aa"),
             selected = "aa")

# Downloadable example FASTA file ----
output$example <- downloadHandler(
    filename = function() {
      paste0("example.fasta", sep = "")
    },
    content = function(file) {
      samp_fil <- readAAStringSet("data/4KU5.fasta")
      writeXStringSet(samp_fil, file)
    }
  )

# Downloadable csv of selected dataset ----
downloadHandler(
      filename = function() {
        paste0("pNPred-", Sys.Date(), ".csv", sep = "")
      },
      content = function(file) {
        rv <- mytab()
        if(nrow(rv) > 0) {
           write_csv(mytab(), file)
        }
      }
    )

mytab <- eventReactive(input$run, {

withProgress(message = 'Processing... Please wait', {

    # Input file
    req(input$file1)
    observeEvent(input$example, {
    })

    tryCatch(
      {
         if(input$seqtype == "aa") {
          sqs <- readAAStringSet(input$file1$datapath)

         }
           else {
          sqs_dna <- readDNAStringSet(input$file1$datapath)
          sqs <- translate(sqs_dna, if.fuzzy.codon = "error")
          if(length(sqs) > 100) {
            ret <- data.frame("Error: please input less than 100 sequences")
            colnames(ret) <- ""
          }
           }

          query_fils <- sapply(1:length(sqs), function(x) {tempfile(pattern = "", fileext = ".fasta")})
          sapply(1:length(sqs), function(x) {writeXStringSet(sqs[x], query_fils[x])})
          
          # Align active site residues and extract
          extract_84_list <- lapply(1:length(sqs), function(x) { extract_12angstrom_aas(query_fils[x]) })
          extract_84_df <- data.frame(matrix(unlist(extract_84_list), nrow = length(extract_84_list), byrow=T), 
                            stringsAsFactors = FALSE)
          
          # Set column names
          colnams84 <- read_csv("data/seqft_colnames.csv", col_names = F) %>%
            pull()
          colnames(extract_84_df) <- colnams84

          # Remove variables with non-zero variance
          which_rem <- read_csv("data/vars_to_remove.csv", col_names = T) %>%
            pull()

          # Combine with substrate features
          hex <- molec_fts[,2:8] 
          subs <- rep(pull(molec_fts[,1]), length(sqs))
          hex_df <- data.frame(cbind(subs, apply(hex, 2, function(x) rep(x, length(sqs)))), stringsAsFactors = F)
          colnames(hex_df) <- c("substrate", paste0("PC", 1:7))
          
          hex_df_sorted <- hex_df[order(hex_df$PC1),] 
          
          ex84_df_scaled <- data.frame(apply(extract_84_df, 2, function(x) rep(x, 15)), stringsAsFactors = F)
          
          test_df <- bind_cols(hex_df_sorted, ex84_df_scaled) %>%
            dplyr::select(-all_of(which_rem), -substrate)
          test_df[,1:7] <- apply(test_df[,1:7], 2, as.numeric)
          
          # Binary classification
          rf_funct_class <- readRDS("data/models/20200324_rf_bin_class_all_data_no_prot_unscaled_optimized.rds")
          rf_funct_class_pred <- predict(rf_funct_class, data = test_df, predict.all = F)
          rf_fc_pred <- rf_funct_class_pred$predictions
          res_fc_prob <- apply(rf_fc_pred, 1, max)
          res_fc <- tibble(colnames(rf_fc_pred)[apply(rf_fc_pred, 1, which.max)])
          
          # Return a data frame of predictions
          pred_df <- data.frame(rep(names(sqs), 15), hex_df_sorted$substrate, res_fc, round(res_fc_prob, 2), stringsAsFactors = F) 

          # Identify active enzymes and predict activity
          rf_reg <- readRDS("data/models/20200401_rf_regression_all_data_no_prot_unscaled_optimized.rds")
          rf_reg <- predict(rf_reg, data = test_df, predict.all = F)
          head(rf_reg$predictions)
          rf_reg_pred <- rf_reg$predictions
          
          reg_df <- data.frame(cbind(pred_df, round(rf_reg_pred, 2)), stringsAsFactors = F)
          colnames(reg_df) <- c("query_name", "subst", "is_active", "pred_prob", "raw_activity")
          
          
          # Show picture of substrate
          reg_df$substrate <- paste0('<img src = "https://github.com/serina-robinson/thiolase-machine-learning/raw/master/shiny_app/www/', ... = reg_df$subst, '.jpeg" style="width: 90%;  height: auto;" >')
          
          reg_df$raw_activity[reg_df$is_active == "N"] <- "No predicted activity"  
          ord_df <- reg_df[order(reg_df$query_name),] %>%
dplyr::mutate(IUPAC_name = case_when(subst == "decanoate" ~ "4-nitrophenyl decanoate",
                    subst == "dodecanoate" ~ "4-nitrophenyl dodecanoate",
                    subst == "7Ph.heptanoate" ~ "4-nitrophenyl 7-phenylheptanoate",
                    subst == "hexanoate" ~ "4-nitrophenyl hexanoate",
                    subst == "heptanoate" ~ "4-nitrophenyl heptanoate",
                    subst == "biotin" ~ "4-nitrophenyl biotin",
                    subst == "heptynoate" ~ "4-nitrophenyl heptynoate",
                    subst == "oxadiazole" ~ "4-nitrophenyl 3-(5-phenyl-1,3,4-oxadiazol-2-yl)propanoate",
                    subst == "dimethyl" ~ "4-nitrophenyl 2,2-dimethylhexanoate",
                    subst == "butoxy" ~ "4-nitrophenyl 2-(2-butoxyethoxy)acetate",
                    subst == "cyclopentyl" ~ "4-nitrophenyl 3-cyclopentylpropanoate",
                    subst == "TMA" ~ "4-nitrophenyl trimethylacetate",
                    subst == "azido" ~ "4-nitrophenyl 6-azidohexanoate",
                    subst == "benzoate" ~ "4-nitrophenyl benzoate",
                    subst == "ClPh.propionate" ~ "4-nitrophenyl-3-(4-chlorophenoxy)propanoate",
                    TRUE ~ "")) %>%
            dplyr::select(query_name, IUPAC_name, substrate, is_active,	pred_prob, raw_activity)
          
          return(ord_df)
},
  error = function(e) {
    # return a safeError if a parsing error occurs
    stop(safeError(e))
    # "Please check input is a valid FASTA file ending in *.faa, *.fa, or *.fasta and beginning with '>'"
  }
  )
})
})

shiny::renderUI(
  HTML(
    mytab() %>%
    mutate(`Query name` = query_name) %>%
    mutate(`Substrate` = substrate) %>%    
    mutate(`IUPAC name` = IUPAC_name) %>%
    mutate(`Active?` = cell_spec(is_active, color = "white",
                               background = case_when(is_active == "Y" & pred_prob >= 0.65 ~ "green",
                                                      is_active == "N" & pred_prob >= 0.65 ~ "darkred",
                                                      pred_prob < 0.65 ~ "grey"))) %>%
    mutate(`Prediction probability` = cell_spec(pred_prob, color = "white",
                               background = case_when(is_active == "Y" & pred_prob >= 0.65 ~ "green",
                                                      is_active == "N" & pred_prob >= 0.65 ~ "darkred",
                                                      pred_prob < 0.65 ~ "grey"))) %>%
    mutate(`Estimated log10 activity ☨` = cell_spec(raw_activity, color = "white",
                                background = case_when(is_active == "Y" & pred_prob >= 0.65 ~ "green",
                                                    is_active == "N" & pred_prob >= 0.65 ~ "darkred",
                                                    pred_prob < 0.65 ~ "grey"))) %>%
    select(-query_name, -pred_prob, -is_active, -raw_activity, -substrate, -IUPAC_name) %>%
    kable(., format = 'html', escape = F) %>%
      row_spec(1:nrow(mytab()), background = "white") %>%
    kable_styling(
        font_size = 15,
        bootstrap_options = c("striped", "hover", "condensed")
       
    )
       
  )
)
```
If the download fails, check that you have first uploaded a FASTA file and clicked the "Run Analysis" button.

☨ Units of activity are log10 of nmol *p*-nitrophenolate produced over the course of one hour by an E. coli BL21 culture with an OD of 1.0 heterologously expressing the enzyme of interest

**Note:** pNPred will make predictions for ALL sequences submitted even if they are not OleA-type enzymes. To limit your input to only thiolases consider first extracting sequences matching the [3-oxoacyl-ACP synthase III (PF08545)](https://pfam.xfam.org/family/ACP_syn_III/) PFAM domain using [HMMscan](https://www.ebi.ac.uk/Tools/hmmer/search/hmmscan)
<br>
<br>
&ast; For nucleotide inputs please ensure the sequence is in-frame with a valid start codon. 
<br>
** Please limit your analysis to 100 sequences or less. If you wish to analyze more sequences, please consider using source code available on [Github](http://github.com/serina-robinson/thiolase-machine-learning/)

<div style="width:20px; height:20px">
[![](www/github_logo.svg)](http://github.com/serina-robinson/thiolase-machine-learning/)
</div>
<br>
<br>

pNP substrates
=====================================
Column {.sidebar data-width=200}
-----------------------------------------------------------------------
The *p*-nitrophenyl esters (pNPs) to the right are the substrates which we have tested to date. We are in the process of expanding our compound libraries and training set to make predictions for new pNP substrates in the future. Stay tuned!

For more information on the synthesis of and 1H-NMR spectra of compounds in this library please see:

Robinson, S.L., Smith, M.D., Richman, J.E., Aukema, K.G., & Wackett, L.P. (2020) Machine learning-based prediction of activity and substrate specificity for OleA enzymes in the thiolase superfamily. *Under review.*

```{r}
# my_list <- list()
# my_list <- lapply(1:nrow(subkey), function(x) { my_list[[x]] <- x })
# names(my_list) <- subkey$`Substrate group name`
# 
# radioButtons("radio_substrate_group", "Substrate group",
#              choices = my_list)
```

Main
-----------------------------------------------------------------------
```{r}

rv <- reactiveValues(data = data.frame(subkey), img = "www/grp1.png")

observeEvent(input$radio_substrate_group, {
    ind <- as.numeric(input$radio_substrate_group)
    rv$data <- data.frame(subkey)[ind,]
    rv$img <- paste0("www/grp", ind, ".png")
})

output$img <- renderImage({
   list(src = rv$img)
}, deleteFile = FALSE)

# renderTable({
#   rv$data
# })

imageOutput("img", width = "80%", height = "60%")
```


Resources
===================================== 
Column {.sidebar}
-----------------------------------------------------------------------

**Welcome to pNPred**
<br>
<br>
*Prediction tool for functional classification and substrate specificity of thiolases towards p-nitrophenyl esters (pNPs)*
<br>
<br>
This tool is maintained by [Dr. Larry Wackett’s lab](https://cbs.umn.edu/wackett-lab/home) at the University of Minnesota.


Inputs
-----------------------------------------------------------------------
**Relevant publications on reactivity of p-nitrophenyl esters with OleA-type thiolases**:

1. Smith, M.D., Robinson, S.L., Molomjamts, M.M., and Wackett, L.P. (2020) Rapid whole cell screen identifying microbial OleA thiolase enzymes involved in olefin and β-lactone biosynthesis. *mBio* 11(2). [doi: 10.1128/mBio.00111-20](https://dx.doi.org/10.1128/mBio.00111-20)

2. Robinson, S.L., Smith, M.D., Richman, J.E., Aukema, K.G., & Wackett, L.P. (2020) Machine learning-based prediction of activity and substrate specificity for OleA enzymes in the thiolase superfamily. *Under review.*

**Relevant publications on OleA structure, mechanism, and diversity**:
  
1. Goblirsch, B.R., Jensen, M.R., Mohamed, F.A., Wackett, L.P., and Wilmot, C.M. (2016) Substrate trapping in crystals of the thiolase OleA identifies three channels that enable long chain olefin biosynthesis. *J. Biol. Chem.* 291(52), 26698-26706. 

2. Goblirsch, B.R., Frias, J.A., Wackett, L.P., and Wilmot, C.M. (2012) Crystal structures of Xanthomonas campestris OleA reveal features that promote head-to-head condensation of two long-chain fatty acids. *Biochemistry* 51(20), 4138-4146. 

3. Frias, J.A., Richman, J.E., Erickson, J.S., and Wackett, L.P. (2011) Purification and characterization of OleA from Xanthomonas campestris and demonstration of a non-decarboxylative Claisen condensation reaction. *J. Biol. Chem.* 286(13), 10930-10938.

4. Sukovich, D.J., Seffernick, J.L., Richman, J.E., Gralnick, J.A., and Wackett L.P. (2010) Widespread head-to-head hydrocarbon biosynthesis in bacteria and role of OleA. *Appl. Environ. Microbiol.* 76(12), 3850-3862.

**Relevant publications on thiolase-family enzymes**:

1. Bonk, B.M., Tarasova, Y., Hicks, M.A., Tidor, B., and Prather, K.L.J. (2018) Rational design of thiolase substrate specificity for metabolic engineering applications. *Biotechnol. Bioeng.*, 115(9), 2167-2182.  

2. Nofiani, R., Philmus, B., Nindita, Y., and Mahmud, T. (2019) 3-Ketoacyl-ACP synthase (KAS) III homologues and their roles in natural product biosynthesis. *MedChemComm* 10(9), 1517-1530.

3. Haapalainen, A.M., Merilainen, G., and Wierenga, R.K. (2006) The thiolase superfamily: condensing enzymes with diverse reaction specificities. *Trends Biochem. Sci.* 31(1), 64-71. 

Database
===================================== 

Inputs {.sidebar}
-----------------------------------------------------------------------
<br>
<br>
Training set database of thiolase enzymes paired with pNP substrates and their activity. 
```{r}

      sliderInput("range", label = "Search by enzyme activity:",
                    min = 0, max = 3,
                    value = c(2.8,2.9),
                    step = 0.05,
      )

       selectizeInput("substrate", "Search by substrate:",
                    choices = tab$Substrate,
                    selected = NULL, multiple = TRUE
      )
      
      selectizeInput("org", "Search by organism:",
                   choices = tab$Organism,
                   selected = NULL, multiple = TRUE
      )
      
```

Outputs
-----------------------------------------------------------------------
```{r}  

# Download button


DT::renderDataTable({
      srch <- tab[tab$Substrate %in% input$substrate,]
      tax <- tab[tab$Organism %in% input$org,]
      minact <- tab[tab$Enzyme_Activity >= input$range[1],]
      maxact <- tab[tab$Enzyme_Activity <= input$range[2],]
      out <- data.frame(rbind(tax, srch, minact), stringsAsFactors = F)
      dedup <- out[!duplicated(out),]

      return(datatable(dedup))
    })

# downloadHandler(
#       filename = function() {
#         paste0("pNPred-", Sys.Date(), ".csv", sep = "")
#       },
#       content = function(file) {
#         write.csv(dedup, file, row.names = FALSE)
#       }
#     )
# 
# downloadButton("downloadData", "Download", class = "download_this", width = '30%')


```   

About
=====================================

Column {.sidebar data-width=200}
-----------------------------------------------------------------------
**Welcome to pNPred**
<br>
<br>
*pNPred uses supervised machine learning approaches to predict substrate specificity of thiolases towards different p-nitrophenyl esters (pNPs).*
<br>
<br>

Main
-----------------------------------------------------------------------
**If you use pNPred in your work, please cite the following reference:**
<br>
<br>
Robinson, S.L., Smith, M.D., Richman, J.E., Aukema, K.G., & Wackett, L.P. (2020) Machine learning-based prediction of activity and substrate specificity for OleA enzymes in the thiolase superfamily. *Under review.*
<br>
<br>
This tool is maintained by [Dr. Larry Wackett’s lab](https://cbs.umn.edu/wackett-lab/home) at the University of Minnesota.
<br>
<br>
The DNA construct synthesis conducted by the U.S. Department of Energy Joint Genome Institute, a DOE Office of Science User Facility, is supported under Contract No. DE-AC02-05CH11231. S.L.R.supported by the National Science Foundation Graduate Research Fellowship under Grant No. 00039202. 
<br>
<br> 
Disclaimer: Any opinions, findings, and conclusions or recommendations expressed in this material are those of the author(s) and do not necessarily reflect the views of the National Science Foundation or the U.S. Department of Energy Joint Genome Institute.
<br>
<br>
**Questions, Comments, or Suggestions?**
Please [contact us](mailto:www.robinsonserinalee@gmail.com)!


